---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Setting up Environment
This short section included libraries used in the analysis and creates a function  
used later in the analysis. I have hidden the library loading messages.
```{r Environment, warning=FALSE, message=FALSE}
# Libraries used in this project
library(readr)
library(lubridate)
library(dplyr)
library(ggplot2)
# Function to convert interval into elapsed minutes
elapsedmin <- function(x){
  e <- ifelse(nchar(x)<=2,paste('00:',x, sep = ""), 
              paste(substr(x,1,nchar(x)-2),":",substr(x, nchar(x)-1, nchar(x)), sep = ""))
  e
}
```

## Loading and preprocessing the data
Load activity data
Preprocess the data frame by adding a variable that identifies weekdays and week ends (used later).  
Also, format the interval data so it can the string value can be later be converted to elapsed time.
```{r Read Data and preprocess, message=FALSE}
# 
d_activity <- read_csv("activity.zip", col_names = TRUE, na = c("NA"))
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
d_activity$wday <- c('weekend', 'weekday')[(weekdays(d_activity$date) %in% weekdays1)+1L]
d_activity$hhmm <- ifelse(nchar(d_activity$interval)<4,paste0(0,d_activity$interval),d_activity$interval)
```


## What is mean total number of steps taken per day?

* Aggregate the daily actiity (total steps / day)  
* Calculate central tendency (median & mean)
* Provide a Histograme of total steps/day

```{r Histogram of the total number of steps taken each day}

a_stepsdays <- aggregate(d_activity$steps, by = list(d_activity$date, d_activity$wday), FUN = sum)
colnames(a_stepsdays) <- c("date", "day", "tsteps")
# Central Tendency
summary(a_stepsdays$tsteps)[3:4]

# Histogram of the total number of steps taken each day ##### 

hist(a_stepsdays$tsteps,
     main = "Histogram of Total Steps/Day",
     xlab = "Total Steps/Day"
     )
```



## What is the average daily activity pattern?

Following plot show daily acitivity averaged over all days (weekdays and weekends).  
Data is steps are averaged over intervals.

```{r Average daily activity}
d_stepsByIntveral <- d_activity %>% group_by(interval) %>% summarize(avgSteps = mean(steps, na.rm = TRUE))
# Create an x-axis based on elapsed minutes of a day (1435 minutes)
d_stepsByIntveral$telapsed <- hm(elapsedmin(d_stepsByIntveral$interval))
d_stepsByIntveral$elapsed <- hour(d_stepsByIntveral$telapsed)*60 + minute(d_stepsByIntveral$telapsed)

plot(d_stepsByIntveral$elapsed, d_stepsByIntveral$avgSteps, type = "l",
     xlab = "Elapsed Minutes",
     ylab = "Average Steps",
     main = "Timeseries of Averages Steps Over the Day")
```



## Imputing missing values
Code to describe and show a strategy for imputing missing data
The next section will show there is not much of a pattern difference between weekday and weekend activities.  Strategy for missing data:  
* Used the entire data median of the data for the 1st day.  
* Used the median of the day before & day after for the other missing data.  

```{r Imputing missing data }
#Determine index of missing data
i_missing <- which(is.na(a_stepsdays$tsteps))
print("List of missing dates:")
a_stepsdays[i_missing,"date"]

e_stepsdays <- a_stepsdays
# Rules
# First day is missing - use median of population
# 
for(i in i_missing){
  ifelse(i==1, e_stepsdays[i,3] <- median(e_stepsdays$tsteps, na.rm = TRUE),
        e_stepsdays[i,3] <- 
          mean(e_stepsdays[(i-1) : (i+1),3], na.rm = TRUE))
}

print("Imputed central tendency")
summary(e_stepsdays$tsteps)[3:4]
print("Central tendency - missing data")
summary(a_stepsdays$tsteps)[3:4]

```



## Are there differences in activity patterns between weekdays and weekends?

```{r Weekday or Weekend activity patterns}
d_weekdayInterval <- filter(d_activity, d_activity$wday == "weekday") %>% 
  group_by(interval) %>% summarize(avgSteps = mean(steps, na.rm = TRUE))
d_weekdayInterval$wday <- "weekday"
d_weekendInterval <- filter(d_activity, d_activity$wday == "weekend") %>% 
  group_by(interval) %>% summarize(avgSteps = mean(steps, na.rm = TRUE))
d_weekendInterval$wday <- "weekend"

g <- ggplot() + geom_line(data = d_weekdayInterval, aes(x = interval, y = avgSteps, color = "Weekday")) +
     geom_line(data = d_weekendInterval, aes(x = interval, y = avgSteps, color = "Weekend")) + 
     facet_grid(wday ~ ., scale = "free_y") + theme_bw() +
     labs(x = "Interal (5-minute interval)", y = "Ave Steps",
       title = "Average Steps Taken across weekdays and weekends", 
       subtitle = "OBS: Activity appears to increase on the weekend") + 
     theme(legend.position = "none")
print(g)

```

